Left-linear rewriting in adhesive categories

### 2024-04-25 concur/fusione.tex
switches

@@

\begin{notation} We  denote any switch of two switchable direct derivations $\dder{D}_0$ and $\dder{D}_1$ along $(i_0, i_1)$ by $S_{i_0, i_1}(\dder{D}_0, \dder{D}_1)$. Moreover, we use $S_{i_0,i_1}(\dder{D}_1)$ and  $S_{i_0,i_1}(\dder{D}_0)$ to denote its components, in such a way that $S_{i_0, i_1}(\dder{D}_0, \dder{D}_1)=S_{i_0,i_1}(\dder{D}_1)\cdot S_{i_0,i_1}(\dder{D}_0)$.
\end{notation}

@@

The first lemma is at the core of globality for independence. It shows that if we have a derivation consisting of three rewriting steps, say $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ where $\dder{D}_0$ and $\dder{D}_1$ are independent, then if we switch $\dder{D}_2$ to the first position, 
obtaining a derivation $\dder{D}_2 \cdot  \dder{D}_0\cdot\dder{D}_1$, the steps $\dder{D}_0$ and $\dder{D}_1$ are still independent.

Instead, if in $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ we have that $\dder{D}_1$ and $\dder{D}_2$ are
independent, and we switch $\dder{D}_0$ to the last position, obtaining a
derivation $\dder{D}_1\cdot \dder{D}_2\cdot  \dder{D}_0$, the steps $\dder{D}_1$ and $\dder{D}_2$ could be no
longer independent, unless an additional condition is
satisfied. Intuitively, this is due to the fact that dealing with
left-linear rewriting systems introduces disjunctive forms of
causality. Hence, if in $\dder{D}_0\cdot \dder{D}_1\cdot \dder{D}_2$ we have that $\dder{D}_0$ and $\dder{D}_1$ are
disjunctive causes of $\dder{D}_2$, at least one of the two is needed to
enable the application of $\dder{D}_2$ and this explain why in
$\dder{D}_1\cdot \dder{D}_2\cdot \dder{D}_0$ we have that $\dder{D}_1$ and $\dder{D}_2$ are no longer
independent. This is exemplified below.

@@

\begin{restatable}[Globality of independence]{lemma}{lemIndepGlobalLeft}
  \label{lem:indep-global-left}
  Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X,\R)$ a well-switching rewriting system.
  For every derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^2$ the following hold
  \begin{enumerate}
  \item
    \label{lem:indep-global-left:1} if $\dder{D}_0\seqind \dder{D}_1$ and there is a switching sequence
    $\der{D}\shift{(1,2)} \der{D}'\shift{(0,1)} \der{D}''$
    then the last two steps of $\der{D}''$ are switchable;

  \item
    \label{lem:indep-global-left:2}
    suppose that $\dder{D}_1 \seqind\dder{D}_2$ with independence pair
    $(\alpha_0, \alpha_1)$ and that, moreover,
    $\dder{D}_0\seqind S_{\alpha_0, \alpha_1}(\dder{D}_2)$, if there
    is a switching sequence
    $\der{D}\shift{(0,1)} \der{E}'\shift{(1,2)} \der{E}''$
    then the
    first two steps of $\der{E}''$ are switchable.
  \end{enumerate}
\end{restatable}
  %

@@

\begin{definition}[Consistent permutation]
	\label{def:permcon}
	Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X, \R)$ 
	a left-linear rewriting system.  Take two derivations $\der{D}:=\{\dder{D}_i\}_{i=0}^n$ and
	$\der{D}':=\{\dder{D}'_i\}_{i=0}^n$ with the same length and with isomorphic sources and targets.
	Let also $r(\der{D})=\{\rho_i\}_{i=0}^n$ and
	$r(\der{D}')=\{\rho'_i\}_{i=0}^n$ be the  associated sequence of rules. 
	A \emph{consistent permutation}
	between $\der{D}$ and
	$\der{D}'$ is a permutation
	$\sigma\colon [0,n]\to [0,n]$ such that, for every $i\in [0,n]$,
	$\rho_i=\rho'_{\sigma(i)}$ and, moreover, there exists a
	\emph{mediating isomorphism}
	$\xi_\sigma\colon \tpro{D} \to \lpro \der{D}' \rpro$ fitting in the
	following diagrams, where $m_i, m'_i, h_i$ and $h'_i$ are, the matches and co-matches of $\dder{D}_i$ and
	$\dder{D}'_i$.
	\[
	\xymatrix@C=30pt{L_i
		\ar[r]^{m_i} \ar[d]_{m'_{\sigma(i)}}& G_i \ar[r]^{\iota_{G_i}}
		&\tpro{D} \ar[d]^{\xi_\sigma} & R_i \ar[r]^{h_i}
		\ar[d]_{h'_{\sigma(i)}}& G_{i+1} \ar[r]^{\iota_{G_{i+1}}}
		&\tpro{D} \ar[d]^{\xi_\sigma} \\G'_{\sigma(i)}
		\ar[rr]_{\iota'_{G'_{\sigma(i)}}}&& \lpro \der{D}' \rpro&
		G'_{\sigma(i)+1} \ar[rr]_{\iota'_{G'_{\sigma(i)+1}}}&& \lpro
		\der{D}' \rpro}
	\]
\end{definition}
We can now turn to the proofs of the results of \Cref{subsec:canonical}.

@@

		\item In addition to the three diagrams above, we have a fourth one, depicted in \Cref{fi:fourth}.
		\begin{figure}[h]
			\centering
			$\xymatrix@C=15pt@R=15pt{&&R_0 \ar@/_.8cm/[ddrr]_(.2){\alpha_0}|(.7)\hole \ar[dr]^{\hspace{-1pt}g_1\circ i_0}&& L_2\ar@/^.8cm/[ddll]^(.2){\alpha_1}  \ar[dl]_{m_2\hspace{-2pt}}\\
				&K_0\ar[dr]^{\hspace{-4pt}q_1\circ v_0}\ar@{>->}[dl]_{l_0} \ar@/_.6cm/[ddrr]|(.36)\hole^(.65){\phi_0}\ar[ur]^{r_0}&& G_2 && K_2 \ar@/^.6cm/[ddll]|(.36)\hole_(.65){\phi_2}\ar[dl]_{k_2}\ar@{>->}[lu]_{l_2} \ar[dr]^{r_2}\\
				L_0 \ar@/_.6cm/[ddrr]_(.2){\gamma_0}|(.31)\hole|(.81)\hole \ar[dr]^(.4){\hspace{-4pt}t_0\circ j_0}|(.6)\hole && Q_1 \ar@{>->}[dl]|(.4)\hole_(.65){s_1}\ar[ur]|(.5)\hole^(.7){t_1}&&D_2 \ar[dr]|(.4)\hole^(.65){g_2} \ar@{>->}[ul]|(.5)\hole_(.65){f_2}&&R_2\ar@/^.6cm/[ddll]^(.2){\gamma_1}|(.31)\hole|(.81)\hole\ar[dl]_(.35){h_2}|(.6)\hole\\&H'_1 &&P_4\ar[dr]_{\lambda_1} \ar@{>->}[dl]^{\lambda_0}\ar[ur]^(.4){\zeta_1}\ar@{>->}[ul]_(.4){\zeta_0}&&G_3\\
				L_2 \ar@/^.6cm/[uurr]^(.2){\alpha_1} \ar[ur]_(.35){\hspace{-2pt}s_1\circ
					\alpha_1}|(.61)\hole&&Q_6
				\ar@{>->}[ul]|(.4)\hole_(.6){\xi_0}\ar[dr]|(.5)\hole^(.65){\xi'_0}
				&&Q_7\ar[ur]|(.4)\hole^(.6){\xi'_1}
				\ar@{>->}[dl]|(.5)\hole_(.65){\xi_1} && R_0 \ar[ul]^(.35){g_2\circ
					\alpha_0}|(.61)\hole\ar@/_.6cm/[uull]_(.2){\alpha_0}\\&K_2
				\ar[ur]_{\lambda_0\circ \phi_2} \ar[dr]_{r_2}
				\ar@{>->}[ul]^{l_2}\ar@/^.6cm/[uurr]_(.65){\phi_2}&&H'_2&& K_0
				\ar[ur]_{r_0} \ar[ul]^{\lambda_1\circ \phi_0} \ar@{>->}[dl]^{l_0}
				\ar@/_.6cm/[uull]^(.65){\phi_0}\\&& R_2
				\ar[ur]_{\hspace{-5pt}\xi_1\circ
					\gamma_1}\ar@/^.8cm/[uurr]^(.2){\gamma_1}&& L_0
				\ar[ul]^{\xi'_0\circ \gamma_0\hspace{-7pt}}
				\ar@/_.8cm/[uull]_(.2){\gamma_0} |(.69)\hole 
			}	$
			\caption{Picture for the proof of \Cref{lem:indep-global-left}: switch of $S_{i_0, i_1}(\dder{D}_0)$ and $\dder{D}_2$.}
			\label{fi:fourth}
		\end{figure}

>>>>

\tikzcdset{
  perm/.style={
    shorten >=-1mm, shorten <=-1mm,
    -,
    % dotted
  }
}

@@

% \begin{notation} We  denote any switch of two switchable direct derivations $\dder{D}_0$ and $\dder{D}_1$ along $(i_0, i_1)$ by $S_{i_0, i_1}(\dder{D}_0, \dder{D}_1)$. Moreover, we use $S_{i_0,i_1}(\dder{D}_1)$ and  $S_{i_0,i_1}(\dder{D}_0)$ to denote its components, in such a way that $S_{i_0, i_1}(\dder{D}_0, \dder{D}_1)=S_{i_0,i_1}(\dder{D}_1)\cdot S_{i_0,i_1}(\dder{D}_0)$.
% \end{notation}

@@

\smallskip

\noindent
\begin{minipage}{.8\textwidth}
  \setlength\parindent{1.5em}
  The first lemma is at the core of globality for independence. It shows that if we have a derivation consisting of three rewriting steps, say $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ where $\dder{D}_0$ and $\dder{D}_1$ are independent, then if we switch $\dder{D}_2$ to the first position, 
obtaining a derivation $\dder{D}_2'' \cdot  \dder{D}_0'\cdot\dder{D}_1'$, as in the diagram on the right where vertical lines represent permutations, the steps $\dder{D}_0'$ and $\dder{D}_1'$ are still independent.
\end{minipage}
\begin{minipage}{.18\textwidth}
\ \hspace{-14mm}
\begin{tikzpicture}[node/.style={draw, minimum size=8mm}]
  \begin{tikzcd}[row sep=2mm, column sep=1mm]
    \dder{D}_0 \arrow[perm,d]  & \dder{D}_1 \arrow[perm,dr] & \dder{D}_2 \arrow[perm,dl]\\
    \dder{D}_0 \arrow[perm,dr]  & \dder{D}_2' \arrow[perm,dl] & \dder{D}_1' \arrow[perm,d]\\
    \dder{D}_2''   & \dder{D}_0' & \dder{D}_1'
  \end{tikzcd}
\end{tikzpicture}
\end{minipage}

\smallskip

\noindent
\begin{minipage}{.8\textwidth}
  \setlength\parindent{1.5em}
  Instead, if in $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ we have that $\dder{D}_1$ and $\dder{D}_2$ are
independent, and we switch $\dder{D}_0$ to the last position, obtaining a
derivation $\dder{D}_1'\cdot \dder{D}_2'\cdot  \dder{D}_0''$, as in the diagram on the right, the steps $\dder{D}_1'$ and $\dder{D}_2'$ could be no
longer independent, unless an additional condition is
satisfied. Intuitively, this is due to the fact that dealing with
left-linear rewriting systems introduces disjunctive forms of
causality. Hence, if in $\dder{D}_0\cdot \dder{D}_1\cdot \dder{D}_2$ we have that $\dder{D}_0$ and $\dder{D}_1$ are
disjunctive causes of $\dder{D}_2$, at least one of the two is needed to
enable the application of $\dder{D}_2$ and this explain why in
$\dder{D}_1'\cdot \dder{D}_2'\cdot \dder{D}_0''$ we have that $\dder{D}_1'$ and $\dder{D}_2'$ are no longer
independent. This is exemplified below.
\end{minipage}
\begin{minipage}{.18\textwidth}
\ \hspace{-14mm}
\begin{tikzpicture}[node/.style={draw, minimum size=8mm}]
     \begin{tikzcd}[row sep=2mm, column sep=1mm]
      \dder{D}_0 \arrow[perm,dr]  & \dder{D}_1 \arrow[perm,dl] & \dder{D}_2 \arrow[perm,d]\\
      \dder{D}_1' \arrow[perm,d]  & \dder{D}_0' \arrow[perm,dr] & \dder{D}_2 \arrow[perm,dl]\\
      \dder{D}_1'   & \dder{D}_2' & \dder{D}_0''
    \end{tikzcd}
  \end{tikzpicture} 
\end{minipage}

@@

\begin{restatable}[Globality of independence]{lemma}{lemIndepGlobalLeft}
  \label{lem:indep-global-left}
  Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X,\R)$ a
  well-switching rewriting system.  Let
  $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ be a three-steps
  derivation.
  \begin{enumerate}
  \item
    \label{lem:indep-global-left:1}
    If $\dder{D}_0$, $\dder{D}_1$ switchable and there is a
    switching sequence
    $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2 \shift{(1,2)}
    \dder{D}_0 \cdot \dder{D}_2' \cdot \dder{D}_1' \shift{(0,1)}
    \dder{D}_2'' \cdot \dder{D}_0' \cdot \dder{D}_1'$ then
    in the last derivation $\dder{D}_0'$ and $\dder{D}_1'$ are switchable;

  \item
    \label{lem:indep-global-left:2}
    Suppose that $\dder{D}_1$,  $\dder{D}_2$ switchable, hence $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2 \shift{(1,2)}
    \dder{D}_0 \cdot \dder{D}_2' \cdot \dder{D}_1'$, and $\dder{D}_0$  $\dder{D}_2'$ switchable.  If there
    is a switching sequence
    $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2 \shift{(0,1)}
    \dder{D}_1' \cdot \dder{D}_0' \cdot \dder{D}_2 \shift{(1,2)}
    \dder{D}_1' \cdot \dder{D}_2' \cdot \dder{D}_0''$ then in the last derivation $\dder{D}_1'$ and $\dder{D}_1'$ are switchable.
  \end{enumerate}
\end{restatable}
  %

@@

\begin{definition}[Consistent permutation]
	\label{def:permcon}
	Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X, \R)$ 
	a left-linear rewriting system.  Take two derivations $\der{D}:=\{\dder{D}_i\}_{i=0}^n$ and
	$\der{D}':=\{\dder{D}'_i\}_{i=0}^n$ with the same length and with isomorphic sources and targets.
	Let also $r(\der{D})=\{\rho_i\}_{i=0}^n$ and
	$r(\der{D}')=\{\rho'_i\}_{i=0}^n$ be the  associated sequence of rules. 
	A \emph{consistent permutation}
	between $\der{D}$ and
	$\der{D}'$ is a permutation
	$\sigma\colon [0,n]\to [0,n]$ such that, for every $i\in [0,n]$,
	$\rho_i=\rho'_{\sigma(i)}$ and, moreover, there exists a
	\emph{mediating isomorphism}
	$\xi_\sigma\colon \tpro{D} \to \lpro \der{D}' \rpro$ fitting in the
	following diagrams, where $m_i, m'_i, h_i$ and $h'_i$ are, the matches and co-matches of $\dder{D}_i$ and
	$\dder{D}'_i$.
	\[
	\xymatrix@C=30pt{L_i
		\ar[r]^{m_i} \ar[d]_{m'_{\sigma(i)}}& G_i \ar[r]^{\iota_{G_i}}
		&\tpro{D} \ar[d]^{\xi_\sigma} & R_i \ar[r]^{h_i}
		\ar[d]_{h'_{\sigma(i)}}& G_{i+1} \ar[r]^{\iota_{G_{i+1}}}
		&\tpro{D} \ar[d]^{\xi_\sigma} \\G'_{\sigma(i)}
		\ar[rr]_{\iota'_{G'_{\sigma(i)}}}&& \lpro \der{D}' \rpro&
		G'_{\sigma(i)+1} \ar[rr]_{\iota'_{G'_{\sigma(i)+1}}}&& \lpro
		\der{D}' \rpro}
	\]
\end{definition}

We can now turn to the proofs of the results of \Cref{subsec:canonical}.

\begin{notation}
  Recall that by \cref{thm:switch_uni} switches are uniquely
  determined, up to abstraction equivalence.
  %
  In the sequel we denote any switch of two switchable direct
  derivations $\dder{D}_0$ and $\dder{D}_1$ along an independence pair
  $(i_0, i_1)$ by $S_{i_0, i_1}(\dder{D}_0, \dder{D}_1)$. Moreover, we
  use $S_{i_0,i_1}(\dder{D}_1)$ and $S_{i_0,i_1}(\dder{D}_0)$ to
  denote its components, in such a way that
  $S_{i_0, i_1}(\dder{D}_0, \dder{D}_1)=S_{i_0,i_1}(\dder{D}_1)\cdot
  S_{i_0,i_1}(\dder{D}_0)$.
\end{notation}

@@

              \item Here we assume that
                $\dder{D}_1$ and $\dder{D}_2$ are switchable  with an
    independence pair $(\alpha_0, \alpha_1)$ and that
    $\dder{D}_0\seqind S_{\alpha_0, \alpha_1}(\dder{D}_2)$.
    % 
                Hence, in addition to the three diagrams above, we have a fourth one, depicted in \Cref{fi:fourth}.
		\begin{figure}[h]
			\centering
			$\xymatrix@C=15pt@R=15pt{&&R_0 \ar@/_.8cm/[ddrr]_(.2){\alpha_0}|(.7)\hole \ar[dr]^{\hspace{-1pt}g_1\circ i_0}&& L_2\ar@/^.8cm/[ddll]^(.2){\alpha_1}  \ar[dl]_{m_2\hspace{-2pt}}\\
				&K_0\ar[dr]^{\hspace{-4pt}q_1\circ v_0}\ar@{>->}[dl]_{l_0} \ar@/_.6cm/[ddrr]|(.36)\hole^(.65){\phi_0}\ar[ur]^{r_0}&& G_2 && K_2 \ar@/^.6cm/[ddll]|(.36)\hole_(.65){\phi_2}\ar[dl]_{k_2}\ar@{>->}[lu]_{l_2} \ar[dr]^{r_2}\\
				L_0 \ar@/_.6cm/[ddrr]_(.2){\gamma_0}|(.31)\hole|(.81)\hole \ar[dr]^(.4){\hspace{-4pt}t_0\circ j_0}|(.6)\hole && Q_1 \ar@{>->}[dl]|(.4)\hole_(.65){s_1}\ar[ur]|(.5)\hole^(.7){t_1}&&D_2 \ar[dr]|(.4)\hole^(.65){g_2} \ar@{>->}[ul]|(.5)\hole_(.65){f_2}&&R_2\ar@/^.6cm/[ddll]^(.2){\gamma_1}|(.31)\hole|(.81)\hole\ar[dl]_(.35){h_2}|(.6)\hole\\&H'_1 &&P_4\ar[dr]_{\lambda_1} \ar@{>->}[dl]^{\lambda_0}\ar[ur]^(.4){\zeta_1}\ar@{>->}[ul]_(.4){\zeta_0}&&G_3\\
				L_2 \ar@/^.6cm/[uurr]^(.2){\alpha_1} \ar[ur]_(.35){\hspace{-2pt}s_1\circ
					\alpha_1}|(.61)\hole&&Q_6
				\ar@{>->}[ul]|(.4)\hole_(.6){\xi_0}\ar[dr]|(.5)\hole^(.65){\xi'_0}
				&&Q_7\ar[ur]|(.4)\hole^(.6){\xi'_1}
				\ar@{>->}[dl]|(.5)\hole_(.65){\xi_1} && R_0 \ar[ul]^(.35){g_2\circ
					\alpha_0}|(.61)\hole\ar@/_.6cm/[uull]_(.2){\alpha_0}\\&K_2
				\ar[ur]_{\lambda_0\circ \phi_2} \ar[dr]_{r_2}
				\ar@{>->}[ul]^{l_2}\ar@/^.6cm/[uurr]_(.65){\phi_2}&&H'_2&& K_0
				\ar[ur]_{r_0} \ar[ul]^{\lambda_1\circ \phi_0} \ar@{>->}[dl]^{l_0}
				\ar@/_.6cm/[uull]^(.65){\phi_0}\\&& R_2
				\ar[ur]_{\hspace{-5pt}\xi_1\circ
					\gamma_1}\ar@/^.8cm/[uurr]^(.2){\gamma_1}&& L_0
				\ar[ul]^{\xi'_0\circ \gamma_0\hspace{-7pt}}
				\ar@/_.8cm/[uull]_(.2){\gamma_0} |(.69)\hole 
			}	$
			\caption{Picture for the proof of \Cref{lem:indep-global-left}: switch of $S_{i_0, i_1}(\dder{D}_0)$ and $\dder{D}_2$.}
			\label{fi:fourth}
		\end{figure}
